---
title: "Exercise 4"
author: "Group C"
date: "17/1/2022"
output:
  pdf_document: default
  html_document: default
---


# Preparing the data
```{r, message=FALSE}
source('exercise4_da_sistemare.R')
library(org.Hs.eg.db)#Homo sapiens database call
library(GO.db)
```

We load the 'dati4.r' file containing selzionati, a vector with the names of genes selected as DE in exercise 3, and totali, a vector with the names of all genes remained after filtering in exercise 3
```{r}
load('dati4.r')
```

We select from the Homo sapiens database the information relative to our genes of interest
```{r, message=FALSE}
#selection of information relative to our genes
gene_GO<-select(org.Hs.eg.db,keys=totali,columns=c('ENTREZID','GOALL'),keytype='SYMBOL')
#eliminate evidenceall column
gene_GO<-gene_GO[,-4]
#select lines that are repeated
doppi<-duplicated(gene_GO[,2:3])
#eliminate repeated lines
gene_GO<-gene_GO[doppi==FALSE,]
```
We obtain a dataframe and here we visualize the first rows
```{r}
head(gene_GO)
```

We add a column to keep track if the gene belongs to the selected ones and we eliminate rows containing NAs
```{r, message=FALSE}
#add a column to track if a gene is part of the selected
SELECTED<-""
gene_GO<-cbind(gene_GO,SELECTED)
gene_GO$SELECTED<-gene_GO$SYMBOL%in%selezionati
#eliminate NAs
gene_GO_NA<-na.omit(gene_GO)
head(gene_GO_NA)
```

We now split the dataframe in 3 different dataframes based on the ONTOLOGYALL denomination, we will run separate analysis for Molecular Function, Cellular Component and Biological Process
```{r}
matriceMF<-gene_GO_NA[gene_GO_NA$ONTOLOGYALL=='MF',]
matriceCC<-gene_GO_NA[gene_GO_NA$ONTOLOGYALL=='CC',]
matriceBP<-gene_GO_NA[gene_GO_NA$ONTOLOGYALL=='BP',]
```


# Enrichement Analysis

## Molecular Function

We now do the step by step analysis for one of the three ontology classes (Molecular Function), for the other ones we will use functions we built and show the results.

Firstly we create a matrix containing, for each unique GO,the total number of occurrences and the number of occurrences in correspondence to the selected genes.

We select the unique GOs
```{r}
GO_present<-duplicated(matriceMF$GOALL)
GO<-matriceMF$GOALL[GO_present==FALSE]
```

And we create the matrix of counts applying the function 'assigned' to each GO.
The function 'assigned' counts the occurrences we want to know by looking in the dataframe
```{r}
count<-sapply(GO,assigned,matriceMF$GOALL,matriceMF$SELECTED)
count<-t(count)
colnames(count)<-c("N_GENES","N_SELECTED")
```

We then remove the GOs for which we have just 1 total count and the ones for which we don't have any selected gene occurrence
```{r}
index_one<-which(count[,1]==1)
FINAL<-count[-index_one,]
FINAL<-FINAL[-which(FINAL[,2]<1),]
```

And we visualize the first rows of the matrix we obtained
```{r}
head(FINAL)
```
Now that we have our counts, we prepare for each G0 term the matrix that we need as input for the fisher exact test.

We count the number of genes remained in our dataframe after the GO selection and the number of selected genes remained.
```{r}
gene_GO_NA<-matriceMF[matriceMF$GOALL%in%rownames(FINAL),]
gene_tot<-duplicated(matriceMF$SYMBOL)
ind_gene<-which(gene_tot==FALSE)
tot_genes<-length(ind_gene)
tot_selected<-length(which(matriceMF$SELECTED[ind_gene]==TRUE))
```

Then we create the input matrix in the following way for each GO remained, we see an example for the first one
```{r}
tot_ac<-FINAL[1,1]
a<- FINAL[1,2]
c<- tot_ac-a
b<- tot_selected-a
d<- tot_genes-c-tot_selected 
#create matrix
mat<-matrix(c(a,c,b,d),2,2)
rownames(mat)<-c('Selected','Non Selected')
colnames(mat)<-c('GO','not GO')
print(mat)
```
We apply to each GO remained the function we built 'fisher_test' that creates the input matrix and then applies the fisher exact test and extracts the pvalue.
We obtain a list with the GOs and corresponding pvalues, we visualize the first elements.
```{r}
p_value<-apply(FINAL,1,fisher_test,tot_selected,tot_genes)
head(p_value)
```
We now have to select the enriched GOs. To to this we decided to find a confidence level alpha for which the FDR is 0.05 and we select all the GOs with p_value less or equal than alpha.

We apply the same procedure we used in exercise 3, so first of all we estimate the value of G0.

```{r}
prima<-G0estimate(p_value,ylim1 = 100,ylim2=400, disegna=FALSE)
```


And we focus on the portion of graph where the estimate gets stable between 0.4 and 0.6
```{r}
stima<-G0estimate(p_value,0.4,0.6,100,400)
```



Then we select the alpha that gives the FDR closest to 0.05
```{r}
coppia<-select_alfa_FDR(p_value,stima,0.05)
names(coppia)<-c('alpha','FDR')
print(coppia)
alfa<-coppia[1]
```
And we select the G0s with pvalue smaller o equal than alpha
```{r}
enrichedMF<-selectgenes(p_value,alfa)
```


We then associate to every GO the corresponding GO term coming from the GO.db database
```{r, message=FALSE}
descriptionMF<-select(GO.db,keys=enrichedMF,columns='TERM',keytype='GOID')
```
We obtain a final matrix containing the enriched GOs and the corresponding GO terms, of which we visualize the first rows
```{r}
head(descriptionMF)
```
We now repeat the analysis for the other two terms.

## Biological Process

For 'Biological Process' we obtain the following.(These are just the first rows)
```{r, message=FALSE, fig.show='hide'}
pvaluesBP<-pvfisher(matriceBP)
enrichedBP<-enrichedGO(pvaluesBP,0.7,0.9,500,3000)
descriptionBP<-select(GO.db,keys=enrichedBP,columns='TERM',keytype='GOID')
head(descriptionBP)
```

## Cellular Component

For 'Cellular Component' we obtain the following.(These are just the first rows)
```{r, message=FALSE,fig.show='hide'}
pvaluesCC<-pvfisher(matriceCC)
enrichedCC<-enrichedGO(pvaluesCC,0.4,0.6,100,400)
descriptionCC<-select(GO.db,keys=enrichedCC,columns='TERM',keytype='GOID')
head(descriptionCC)
```





